A general and efficient representation of Ancestral Recombination Graphs

### 2022-09-21 paper.tex
Remove description of W&H node u as a MRCA
I think the only common ancestors that are present in the simplified TS will be o and v
Some trivial edits and comments

% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
An extensive body of work has developed in
studying the coalescent with recombination
% Can we add a couple of the most well known CwR examples here
and other related
graph-valued stochastic processes from a mathematical perspective.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection instead of recombination.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode
desired patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
% Could mention the Ancestral Gene Transfer graph here too, within
% something like "Similarly the Ancestral Gene Transfer Graph [cite]
% and [other ancestral graphs, if they exist?] model genealogical
% processes in bacteria as a graph.

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the
stochastic process generating genealogies in the presence of
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent},
as well as, more commonly, the concrete realisations of the
ancestry~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

@@

% The eARG encoding consists of two types of event
% (common ancestor and recombination) but nature is not so parsimonious
% and there are many other processes that will affect the
% ancestry of samples.
% For example, gene conversion between homologous
% chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
% evolutionary force in many species~\cite[e.g.][]{yang2012great}.
% % Some refs for where people are inferring ancestry in the
% % presence of GC? Isn't this what ClonalFrame etc are all about?
% Similarly, the eARG encoding cannot easily represent multiple
% recombinations on a chromosome, or the transmission of multiple
% chromosomes.
% What we can infer about ancestral histories and
% processes should not be limited by the data structure that we
% use to represent such ancestry. The eARG is simply one way
% to \emph{encode} such history, and, as we have shown,
% one that has significant limitations of expressivity
% derived from its historical origins in the idealised
% coalescent with recombination model.

@@

% \section*{Genome ARGs}
There is no fundamental reason, however, that the data structure we use
to describe patterns of genetic inheritance must be in terms
of specific causal events. As we demonstrate in subsequent sections,
there are significant advantages to using
a slightly different formulation, which
we call a ``genome ARG'' (gARG).
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance). To recover the local tree for a genomic
position $x$, we traverse the graph rootwards from the leaves
(as in an eARG) and at a particular node, we follow the outbound
edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure
in the gARG encoding),
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
The difference between eARGs and gARGs are subtle, and whether
we choose to interpret nodes as events or genomes
is often of little practical importance.
However, as we show in the remainder of this article,
there are substantial practical gains
from using the gARG encoding, allowing differential
levels of information about the recombination process to be
represented in a computionally efficient way.
The key to this is not so much in how we interpret nodes
(see Appendix XXX), but specifically in how
we represent the process of genetic inheritance.
The critical distinction is that
the gARG encoding allows us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, something that
cannot be done by storing recombination breakpoints.

@@

Fig~\ref{fig-ancestry-resolution}A is an instructive example
from the literature \citep{wiuf1999recombination} and illustrates
how a classical eARG can be described as a gARG with local edge annotations
(directly corresponding to the eARG description), and how we can
then derive the ancestry resolved edge annotations.
It is important to note, however, that the example is a sample from
the so-called ``big'' ARG \emph{process} (see Appendix XXX)
and has several features that we are unlikely to simulate or infer in practice.
For example, by definition the samples contain no genetic information about
node \textsf{j} and so it is unlikey to be inferred.
Likewise, is is unlikely we would have information to infer the ``grand MRCA''
\textsf{w} as all regions of the genome have already reached a
common ancestor separately
(in nodes \textsf{o}, \textsf{u}  and \textsf{v}).

@@

These alternatives of recording the direct inheritance of genetic material from
parent to child vs the propagation of genomic intervals ancestral to the sample
are analogous to running forwards-time vs backwards-time simulations.
Indeed, the process of taking a graph with edges annotated with direct parent-child
inheritance (Fig~\ref{fig-ancestry-resolution}A) and producing the equivalent
``ancestry resolved'' graph in which only those genomic intervals that are
ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}B)
is the same key algorithm underlying recent advances in forwards-time simulation.
\cite{kelleher2018efficient} developed a method in which forwards-time
simulations are performed by recording all parent-child genetic relationships
(i.e.~annotating the edges in the simulated pedigree with the genomic
intervals inherited by a child from a parent) and periodically ``simplifying''
this structure to retain only ancestral relationships that are relevant
to the ancestry of the current population. In many situations, recording
the full genetic ancestry of the population in this way is orders of magnitude faster
than the alternative of tracking
sequences~\citep{kelleher2018efficient,haller2018tree}.

@@

\section*{Simplification}\label{ARG_simplification}
ARG simplification~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy,
retaining the ``meaningful'' topology within local trees
and maintaining the identity of nodes across different trees.
The idea of meaningful local tree topologies is based on the
presence of ``unary'' nodes: a tree node that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right.
For example, consider the path above node
\textsf{a} in the highlighted tree (covering positions 60 to 70)
in Fig.~\ref{fig-simplification}A.
Before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes.
This reflects the
passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For most purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative nodes from the local trees by modifying the gARG.

@@

% If we want to use the coalescent, then really the assumptions
% need to make sense.
% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
Another core assumption of the coalescent model is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.

>>>>

% Mathematical treatment of the CwR and ARGs (and ARG-like stochastic processes)
An extensive body of work has been developed from
studying the coalescent with recombination
% Can we add a couple of the most well known CwR examples here
and other related
graph-valued stochastic processes from a mathematical perspective.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection instead of recombination.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode
desired patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
% Could mention the Ancestral Gene Transfer graph here too, within
% something like "Similarly the Ancestral Gene Transfer Graph [cite]
% and [other ancestral graphs, if they exist?] model genealogical
% processes in bacteria as a graph.

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the
stochastic process generating genealogies in the presence of
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent},
as well as, more commonly, the concrete realisation of ancestry from a
process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

@@

% The eARG encoding consists of two types of event
% (common ancestor and recombination) but nature is not so parsimonious
% and there are many other processes that will affect the
% ancestry of samples.
% For example, gene conversion between homologous
% chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
% evolutionary force in many species~\cite[e.g.][]{yang2012great}.
% % Some refs for where people are inferring ancestry in the
% % presence of GC? Isn't this what ClonalFrame etc are all about?
% Similarly, the eARG encoding cannot easily represent multiple
% recombinations on a chromosome, or the transmission of multiple
% chromosomes.
% What we can infer about ancestral histories and
% processes should not be limited by the data structure that we
% use to represent such ancestry. The eARG is simply one way
% to \emph{encode} such history, and, as we have shown,
% one that has significant limitations of expressivity
% derived from its historical origins in the idealised model of
% coalescent with recombination.

@@

% \section*{Genome ARGs}
There is no fundamental reason, however, that the data structure we use
to describe patterns of genetic inheritance must be in terms
of specific causal events. As we demonstrate in subsequent sections,
there are significant advantages to using
a slightly different formulation, which
we call a ``genome ARG'' (gARG).
First, let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a gARG is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance). To recover the local tree for a genomic
position $x$, we traverse the graph rootwards from the leaves
(as in an eARG) and at a particular node, we follow the outbound
edge associated with the genomic interval containing $x$.
% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree}
shows how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
In this diploid setting
(there is no limitation on mating system, ploidy, or age structure
in the gARG encoding),
a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty using this encoding.

@@

% The reader may arrive at the end of this section going, er, so what?
% Try to tackle this head on by getting to the heart of the distinction.
The difference between eARGs and gARGs are subtle, and whether
we choose to interpret nodes as events or genomes
might seem of little practical importance.
However, as we show in the remainder of this article,
there are substantial practical gains
from using the gARG encoding, allowing differential
levels of information about the recombination process to be
represented in a computionally efficient way.
The key to this is not so much in how we interpret nodes
(see Appendix XXX), but specifically in how
we represent the process of genetic inheritance.
The critical distinction is that
the gARG encoding allows us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, something that
cannot be done by storing recombination breakpoints.

@@

Fig~\ref{fig-ancestry-resolution}A is an instructive example
from the literature \citep{wiuf1999recombination} and illustrates
how a classical eARG can be described as a gARG with local edge annotations
(directly corresponding to the eARG description), and how we can
then derive the ancestry resolved edge annotations.
It is important to note, however, that the example is a sample from
the so-called ``big'' ARG \emph{process} (see Appendix XXX)
and has several features that we are unlikely to simulate or infer in practice.
For example, by definition the samples contain no genetic information about
node \textsf{j} and so it is unlikey to be inferred.
Likewise, is is unlikely we would have information to infer the ``grand MRCA''
\textsf{w} as all regions of the genome have already reached a
common ancestor separately
(in nodes \textsf{o} and \textsf{v}).

@@

These alternatives of recording the direct inheritance of genetic material from
parent to child vs the propagation of genomic intervals ancestral to the sample
are analogous to running forwards-in-time vs backwards-in-time simulations.
Indeed, the process of taking a graph with edges annotated with direct parent-child
inheritance (Fig~\ref{fig-ancestry-resolution}A) and producing the equivalent
``ancestry resolved'' graph in which only those genomic intervals that are
ancestral to the sample are retained (Fig~\ref{fig-ancestry-resolution}B)
is the same key algorithm underlying recent advances in forwards-in-time simulation.
\cite{kelleher2018efficient} developed a method in which forwards-in-time
simulations are performed by recording all parent-child genetic relationships
(i.e.~annotating the edges in the simulated pedigree with the genomic
intervals inherited by a child from a parent) and periodically ``simplifying''
this structure to retain only ancestral relationships that are relevant
to the ancestry of the current population. In many situations, recording
the full genetic ancestry of the current population in this way is orders of magnitude faster
than the alternative of tracking
sequences~\citep{kelleher2018efficient,haller2018tree}.

@@

\section*{Simplification}\label{ARG_simplification}
% I think a sentence is needed here on why "so much" text is devoted to simplification.
% How does simplification connect to eARG vs gARG and practical benefits of gARG over eARG
% and why do you now talk about unary nodes and a bit polytomies? Hold my hand here please;)
ARG simplification~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy,
retaining the ``meaningful'' topology within local trees
and maintaining the identity of nodes across different trees.
The idea of meaningful local tree topologies is based on the
presence of ``unary'' nodes: a tree node that has exactly
% should presence above be absence?
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a
(backwards-in-time) Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right.
For example, consider the path above node
\textsf{a} in the highlighted tree (covering positions 60 to 70)
in Fig.~\ref{fig-simplification}A.
Before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through \textsf{e}, \textsf{j}, \textsf{m}, \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes.
This reflects the
passage of this span of genome through the ARG towards the root.
Such unary nodes in a local tree are not usually considered important;
% because they provide little information and are not
% detectable by information from any given individual tree.
nodes usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For most purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be equivalent to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative nodes from the local trees by modifying the gARG.

@@

% If we want to use the coalescent, then really the assumptions
% need to make sense.
% n << Ne is a silly assumption in today's datasets
A core assumption of the coalescent is that the sample size $n$
is much less than the effective population size, $N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
Agricultural datasets are even more extreme.
For example,
the US dairy cattle database alone currently comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome lineages present in today's US Holstein dairy breed~\citep{Yue2015}.
Simlarly, the 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
In another example, genome and SNP array data for 440,610 individuals within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}
were combined to infer recombination in pigs~\citep{RosFreixedes2022}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
Another core assumption of the coalescent model is that the genome (or
at least the region under study) is short enough that the number of extant
lineages remains much smaller than $N_e$ at all times. Whole
genome sequences have been available for model organisms
for over a decade now, % True? Citation?
and indeed complete chromosome-level assemblies are possible
in humans~\citep{miga2020telomere}.
Projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
